---
title: "ArchR ATAC RDS → AnnData Export"
subtitle: "For Lai et al. Nature 2024 skeletal muscle aging snATAC-seq (hg38)"
author: "Akanksha + Claude"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(Matrix)
library(SummarizedExperiment)
```

# 1. Configuration

```{r config}
# ── Update these ──
rds_path   <- "/ix/djishnu/Akanksha/datasets/SKM_multimodal_ageing/objects/atac_fiber/Fiber_snATAC_65380.RDS"
export_dir <- "/ix/djishnu/Akanksha/datasets/SKM_multimodal_ageing/objects"

dir.create(export_dir, recursive = TRUE, showWarnings = FALSE)
```

# 2. Load and identify what's inside

ArchR pipelines can save different things as RDS: the full ArchR project, 
a SummarizedExperiment (from `getMatrixFromProject`), or a Seurat object 
(from the CCA integration step). This chunk figures out which one you have.

```{r load}
obj <- readRDS(rds_path)

obj_class <- class(obj)
cat("Class:", obj_class, "\n")

if (any(grepl("ArchRProject", obj_class))) {
  cat("\n→ ArchRProject object\n")
  cat("  NOTE: Arrow files must be accessible at their original paths.\n")
  cat("  If moved, call obj@arrowFiles and update paths.\n")
} else if (inherits(obj, "RangedSummarizedExperiment") || inherits(obj, "SummarizedExperiment")) {
  cat("\n→ SummarizedExperiment (likely from getMatrixFromProject)\n")
  cat("  Dimensions:", dim(obj), "\n")
  cat("  Assay names:", paste(names(SummarizedExperiment::assays(obj)), collapse = ", "), "\n")
} else {
  cat("\n→ Unknown type\n")
  str(obj, max.level = 2)
}
```
``` {r}
cat("Value range:", range(assay(obj)@x), "\n")
cat("Integer-valued:", all(assay(obj)@x == floor(assay(obj)@x)), "\n")
```
# 3. Extract peak counts

**Pick the section (3a / 3b / 3c) matching Step 2 output.** Set `eval=TRUE` 
on that chunk, leave others `eval=FALSE`.

## 3a. ArchRProject → extract PeakMatrix

This is the most likely case. Requires ArchR installed and Arrow files 
accessible on disk.

```{r extract_archr, eval=FALSE}
library(ArchR)

# If arrow files were moved, fix paths first:
# obj@arrowFiles <- gsub("/old/path/", "/new/path/", obj@arrowFiles)

cat("Arrow files:\n")
print(obj@arrowFiles)
cat("\nAvailable matrices:\n")
print(getAvailableMatrices(obj))

# ── Peak matrix ──
peak_se <- getMatrixFromProject(obj, useMatrix = "PeakMatrix")
peak_counts <- assay(peak_se)

# Build peak names from GenomicRanges
gr <- rowRanges(peak_se)
peak_names <- paste0(seqnames(gr), "-", start(gr), "-", end(gr))
rownames(peak_counts) <- peak_names

cat("Peak matrix:", nrow(peak_counts), "peaks x", ncol(peak_counts), "cells\n")
cat("Peak name format:", head(peak_names, 3), "\n")

# ── Cell metadata ──
meta <- as.data.frame(getCellColData(obj))
cat("Metadata columns:", paste(colnames(meta), collapse = ", "), "\n")

# ── LSI embedding (from addIterativeLSI) ──
lsi_emb <- getReducedDims(obj, reducedDims = "IterativeLSI")
cat("LSI dims:", ncol(lsi_emb), "\n")

# ── Harmony embedding (if present) ──
harmony_emb <- tryCatch(
  getReducedDims(obj, reducedDims = "Harmony"),
  error = function(e) NULL
)
if (!is.null(harmony_emb)) cat("Harmony dims:", ncol(harmony_emb), "\n")

# ── UMAP (if present) ──
umap_emb <- tryCatch(
  getEmbedding(obj, embedding = "UMAP"),
  error = function(e) NULL
)
if (!is.null(umap_emb)) cat("UMAP dims:", ncol(umap_emb), "\n")

# ── Optional: GeneScoreMatrix (useful for MultiVI) ──
# gene_se <- getMatrixFromProject(obj, useMatrix = "GeneScoreMatrix")
# gene_scores <- assay(gene_se)
# gene_names <- rowData(gene_se)$name
```

## 3b. SummarizedExperiment (pre-exported peak matrix)

```{r extract_se, eval=FALSE}
library(SummarizedExperiment)

peak_counts <- assay(obj)

if (inherits(obj, "RangedSummarizedExperiment")) {
  gr <- rowRanges(obj)
  peak_names <- paste0(seqnames(gr), "-", start(gr), "-", end(gr))
  rownames(peak_counts) <- peak_names
}

cat("Peak matrix:", nrow(peak_counts), "peaks x", ncol(peak_counts), "cells\n")
cat("Peak name format:", head(rownames(peak_counts), 3), "\n")

meta <- as.data.frame(colData(obj))
lsi_emb <- NULL; harmony_emb <- NULL; umap_emb <- NULL
```
````{r extract_se, eval=TRUE}
library(SummarizedExperiment)

peak_counts <- assay(obj)

if (inherits(obj, "RangedSummarizedExperiment")) {
  gr <- rowRanges(obj)
  peak_names <- paste0(seqnames(gr), "-", start(gr), "-", end(gr))
  rownames(peak_counts) <- peak_names
}

cat("Peak matrix:", nrow(peak_counts), "peaks x", ncol(peak_counts), "cells\n")

meta <- as.data.frame(colData(obj))

# ── Try to recover embeddings from colData ────────────────────────────────────
# ArchR sometimes stores LSI/UMAP dims as colData columns
lsi_cols     <- grep("^LSI_|^IterativeLSI", colnames(meta), value = TRUE)
harmony_cols <- grep("^Harmony_|^harmony_", colnames(meta), value = TRUE)
umap_cols    <- grep("^UMAP_|^harmony_UMAP", colnames(meta), value = TRUE)

lsi_emb     <- if (length(lsi_cols) > 0)     as.matrix(meta[, lsi_cols])     else NULL
harmony_emb <- if (length(harmony_cols) > 0) as.matrix(meta[, harmony_cols]) else NULL
umap_emb    <- if (length(umap_cols) > 0)    as.matrix(meta[, umap_cols])    else NULL

cat("LSI in colData:", !is.null(lsi_emb), "\n")
cat("Harmony in colData:", !is.null(harmony_emb), "\n")
cat("UMAP in colData:", !is.null(umap_emb), "\n")
````
```{r export umap}
umap_cols <- grep("^UMAP_|^harmony_UMAP", colnames(meta), value = TRUE)
cat("UMAP columns found:", umap_cols, "\n")  # show exact names
umap_emb <- as.matrix(meta[, umap_cols])
write.table(umap_emb, file.path(export_dir, "embedding_UMAP.tsv"),
            sep = "\t", quote = FALSE, col.names = NA)
cat("✓ UMAP exported:", nrow(umap_emb), "cells x", ncol(umap_emb), "dims\n")
```

## 3c. Seurat object (from CCA integration step)

```{r extract_seurat, eval=FALSE}
library(Seurat)

cat("Assays:", paste(names(obj@assays), collapse = ", "), "\n")

# Find the ATAC assay
atac_assay <- intersect(c("peaks", "ATAC", "Peaks", "chromatin"), names(obj@assays))[1]
if (is.na(atac_assay)) {
  cat("Available assays:", names(obj@assays), "\n")
  stop("Set atac_assay manually")
}

peak_counts <- GetAssayData(obj, assay = atac_assay, slot = "counts")
cat("Peak matrix:", nrow(peak_counts), "peaks x", ncol(peak_counts), "cells\n")
cat("Peak name format:", head(rownames(peak_counts), 3), "\n")

meta <- obj@meta.data
lsi_emb <- tryCatch(Embeddings(obj, "lsi"), error = function(e) NULL)
harmony_emb <- tryCatch(Embeddings(obj, "harmony"), error = function(e) NULL)
umap_emb <- tryCatch(Embeddings(obj, "umap"), error = function(e) NULL)
```

# 4. Sanity checks

```{r sanity}
if (!inherits(peak_counts, "dgCMatrix")) {
  peak_counts <- as(peak_counts, "dgCMatrix")
}

cat("── Sanity checks ──\n")
cat("Dimensions:", nrow(peak_counts), "peaks x", ncol(peak_counts), "cells\n")
cat("Nonzeros:", nnzero(peak_counts), "\n")
cat("Sparsity:", round((1 - nnzero(peak_counts) / prod(dim(peak_counts))) * 100, 1), "%\n")
cat("Value range:", range(peak_counts@x), "\n")
cat("Integer-valued:", all(peak_counts@x == floor(peak_counts@x)), "\n")

cat("\nFirst 5 peaks:\n"); print(head(rownames(peak_counts), 5))
cat("\nFirst 5 barcodes:\n"); print(head(colnames(peak_counts), 5))
cat("\nMetadata:", nrow(meta), "cells,", ncol(meta), "cols\n")
if (ncol(meta) > 0) cat("Columns:", paste(colnames(meta), collapse = ", "), "\n")
```

# 5. Standardize peak names for scvi-tools

scvi-tools / MultiVI expects `chr1-1000-2000` (all hyphens). ArchR 
GenomicRanges sometimes produce `chr1:1000-2000`.

```{r fix_peaks}
pn <- rownames(peak_counts)
if (any(grepl(":", pn))) {
  cat("Fixing colon → hyphen in peak names...\n")
  pn <- gsub(":", "-", pn)
  rownames(peak_counts) <- pn
}
cat("Format OK:", head(pn, 3), "\n")
```

# 6. Export peak counts

```{r export_counts}
writeMM(peak_counts, file.path(export_dir, "atac_counts.mtx"))
writeLines(rownames(peak_counts), file.path(export_dir, "peaks.tsv"))
writeLines(colnames(peak_counts), file.path(export_dir, "barcodes.tsv"))
cat("✓ Peak counts exported\n")
```

# 7. Export metadata

```{r export_meta}
# ArchR barcodes have format "sampleName#BARCODE-1" — preserve as-is
write.csv(meta, file.path(export_dir, "metadata.csv"), row.names = TRUE)
cat("✓ Metadata:", ncol(meta), "columns\n")

# Show transferred cell type labels if present
label_cols <- grep("cell.?type|cluster|label|predict|annot", colnames(meta),
                   ignore.case = TRUE, value = TRUE)
if (length(label_cols) > 0) {
  cat("\nCell type columns found:", paste(label_cols, collapse = ", "), "\n")
  for (lc in label_cols) {
    cat("\n", lc, ":\n")
    print(head(sort(table(meta[[lc]]), decreasing = TRUE), 10))
  }
}
```

# 8. Export embeddings

```{r export_emb}
# LSI
if (exists("lsi_emb") && !is.null(lsi_emb)) {
  write.table(lsi_emb, file.path(export_dir, "embedding_LSI.tsv"),
              sep = "\t", quote = FALSE, col.names = NA)
  cat("✓ LSI:", ncol(lsi_emb), "dims\n")
}

# Harmony
if (exists("harmony_emb") && !is.null(harmony_emb)) {
  write.table(harmony_emb, file.path(export_dir, "embedding_harmony.tsv"),
              sep = "\t", quote = FALSE, col.names = NA)
  cat("✓ Harmony:", ncol(harmony_emb), "dims\n")
}

# UMAP
if (exists("umap_emb") && !is.null(umap_emb)) {
  write.table(umap_emb, file.path(export_dir, "embedding_UMAP.tsv"),
              sep = "\t", quote = FALSE, col.names = NA)
  cat("✓ UMAP:", ncol(umap_emb), "dims\n")
}
```

# 9. Export summary

```{r summary}
flist <- list.files(export_dir)
cat("── Exported to", export_dir, "──\n")
for (f in flist) {
  sz <- round(file.info(file.path(export_dir, f))$size / 1e6, 2)
  cat("  ", f, "(", sz, "MB)\n")
}
```

